---
title: "EnrichmentMap Example"
output:
  github_document: default
  html_notebook: default
---
  
```{r knitrSetup, include=FALSE}
library(knitr)
opts_chunk$set(out.extra='style="display:block; margin: auto"', fig.align="center", tidy=FALSE)
verbose <- FALSE
```

# Introduction 
This repository has scripts to use Pathway Commons to generate [EnrichmentMaps](https://www.ncbi.nlm.nih.gov/pubmed/21085593) using the [clusterProfiler](https://bioconductor.org/packages/release/bioc/html/clusterProfiler.html) with some modification. 

## Data 
Uses gene lists from the paper [Jordan NV et al., HER2 expression identifies dynamic functional states within circulating breast cancer cells, Nature, 2016](https://www.ncbi.nlm.nih.gov/pubmed/27556950) developed on a single-cell dataset derived from circulating tumour cells from 19 women with ER+/HER2- primary tumours.

# Load libraries 
```{r}
library(clusterProfiler)
library(paxtoolsr)
library(org.Hs.eg.db)
library(tidygraph)
library(tidytext)
library(readxl)
library(magrittr)

source("modified_clusterProfiler_functions.R")

# Return variable as string 
# From: https://stackoverflow.com/questions/14577412/how-to-convert-variable-object-name-into-string
curVar <- function(x) { deparse(substitute(x)) }
```


# Load Data 
```{r}
data(stop_words)

data(geneList, package="DOSE")
dose_geneList <- geneList

upErbb2Lo <- read_xlsx("nature19328-s5.xlsx", skip = 2, sheet = "Up in ERBB2 low")
upErbb2Hi <- read_xlsx("nature19328-s5.xlsx", skip = 2, sheet = "Up in ERBB2 high")
```

# Convert Pathway Commons GMT to ClusterProfiler format 
```{r}
tmpFile <- "PathwayCommons11.All.hgnc.gmt.gz"
if(!file.exists(tmpFile)) {
  download.file("https://www.pathwaycommons.org/archives/PC2/v11/PathwayCommons11.All.hgnc.gmt.gz", tmpFile)
} 
gmt <- readGmt(tmpFile, returnInfo = TRUE)  
```

# Parameters
```{r}
maxTags <- 5
```

# Convert Pathway Commons GMT to clusterProfiler structure
```{r}
geneSetList <- lapply(seq_along(gmt), function(x, n, i) { 
  tmp <- x[[i]]
  data.frame(id=n[i], name=tmp[["name"]], gene=tmp[["geneSet"]], stringsAsFactors=FALSE)
}, x=gmt, n=names(gmt))

tmp <- do.call("rbind", geneSetList)
rownames(tmp) <- 1:nrow(tmp) # For convenience 

pc2gene <- tmp[, c("id", "gene")]
pc2name <- tmp[, c("id", "name")]
```

# Run Enrichment Analysis 
```{r}
enrichOutput_upErbb2Lo <- enricher(upErbb2Lo$Symbol, pvalueCutoff=0.05, minGSSize=10, maxGSSize=500, TERM2GENE=pc2gene, TERM2NAME=pc2name)
enrichOutput_upErbb2Lo@result

enrichOutput_upErbb2Hi <- enricher(upErbb2Hi$Symbol, pvalueCutoff=0.05, minGSSize=10, maxGSSize=500, TERM2GENE=pc2gene, TERM2NAME=pc2name)
enrichOutput_upErbb2Hi@result
```

# Plots 
## Plot Default EnrichmentMap  
```{r}
em <- emapplot(enrichOutput_upErbb2Lo, showCategory=20)
em
```

## Modified EnrichmentMap Plot
### Extract graph from enricher output
```{r}
g_upErbb2Lo <- emapplotModGraph(enrichOutput_upErbb2Lo, showCategory=20)
g_upErbb2Hi <- emapplotModGraph(enrichOutput_upErbb2Hi, showCategory=20)
#plot(g)
```

### Merge graphs
```{r}
g_combined <- g_upErbb2Lo %u% g_upErbb2Hi

# Merge attributes
V(g_combined)$color <- ifelse(is.na(V(g_combined)$color_1), 
                              -1*log10(V(g_combined)$color_2), 
                              log10(V(g_combined)$color_1))

V(g_combined)$size <- ifelse(is.na(V(g_combined)$size_1), V(g_combined)$size_2, V(g_combined)$size_1)
E(g_combined)$width <- ifelse(is.na(E(g_combined)$width_1), E(g_combined)$width_2, E(g_combined)$width_1)

# vertex_attr(g_combined, "color")
# vertex_attr(g_combined, "size")
# edge_attr(g_combined, "width")
```

### Modify graph 
```{r}
g <- g_combined

tmp <- components(g) 
gComponents <- data.frame(name=names(tmp$membership), component=as.numeric(tmp$membership))

#tmp <- cluster_fast_greedy(g) 
#gComponents <- data.frame(name=tmp$names, component=as.numeric(tmp$membership))

g1 <- as_tbl_graph(g)
nodes <- g1 %>% activate(nodes) %>% as_tibble
edges <- g1 %>% activate(edges) %>% as_tibble
nodeOrder <- nodes$name
#limits <- c(min(nodes$color), max(nodes$color))
tmp <- max(abs(nodes$color), abs(nodes$color))
limits <- c(-tmp, tmp)

tmpNodes <- merge(nodes, gComponents, by="name", all.x=TRUE)
nodesOrdered <- tmpNodes[match(nodeOrder, tmpNodes$name),]

table(nodesOrdered$component)

g2 <- tbl_graph(
  nodes = nodesOrdered, 
  edges = edges, 
  directed = FALSE
)

tags <- sapply(1:max(nodesOrdered$component), function(i) {
  #i <- 2
  tmp <- nodesOrdered[nodesOrdered$component == i, "name"]
  #if(length(tmp) == 1) { 
  #  tags <- tmp 
  #} else {
    #tmp <- gsub("[[:punct:]]", " ", tmp)
    tmp <- gsub("[^[:alnum:]-]", " ", tmp)

    tmp <- gsub("  ", " ", tmp)
    tmp <- tmp %>% tolower
    
    stopWords <- c(stop_words$word, 
                   "signaling", "interactions", "cells", "cell", "mediated", 
                   "involved", "genes")
    
    a <- strsplit(tmp, " ") %>% unlist %>% unique 
    b <- a[!(a %in% stopWords)]
    b <- b[nchar(b) >= 3]
    
    c <- sapply(b, function(x) {
      grepl(x, tmp) %>% which %>% length 
    }) %>% sort(decreasing = TRUE)
    
    if(max(c) > 1) {
      d <- c[c > 1] %>% sort(decreasing = TRUE)
    } else {
      d <- c %>% sort(decreasing = TRUE)
    }
    
    maxTags <- ifelse(length(d) > maxTags, maxTags, length(d))
    tags <- paste(names(d)[1:maxTags], collapse = ", ")
  #}
  
  tags
})

tags <- data.frame(component=1:max(nodesOrdered$component), tag=paste0("Tags: ", tags), stringsAsFactors = FALSE)
```

### Plot Modified EnrichmentMap
```{r, fig.width=15, fig.height=15}
emapplotModPlot(g2, limits=limits, tags=tags)
```